Supplement to ``{\mytitle}''

### 2024-02-13 si/calibrateMod3.Rnw
refactored fitting model 3 without hurricane parameters. INCOMPLETE

<<CalibrateOrig3>>=
# Create vectors for the unit and shared parameters
unit_specific_names <- c("betaB", "foi_add")
shared_param_names <- c(
  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
  "epsilon", "k"
)
est_param_names <- c(
  unit_specific_names, shared_param_names
)

@@

haiti3_sub_fit <- bake(
  file = "../model3/si/subset_noparams.rds", {
    fit_haiti3(
      nsearches = 2L,
      search1 = SEARCH1,
      search2 = SEARCH2,
      search_rho = FALSE,
      search_gamma = FALSE,
      search_Iinit = FALSE,
      search_hur = FALSE
    )
  }
)

@@

# Perform block partical filter with fitted model, to get conditional
# log-likelihoods.
bpf_sub3_out <- bake(
  file = "../model3/si/bpf_sub_cond_LL.rds", {

@@

# This result was used in the manuscript, so it doesn't need to be recalculated.
h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")

# Save best_m as index for the model parameters with the best likelihood.
# Then, save the model parameter values and the likelihood evaluations
best_m <- which.max(h3_final_results$search2$logLiks$logLik)
best_params <- h3_final_results$search2$params[best_m, ]
all_liks <- as.data.frame(h3_final_results$search2$likMat)
all_liks$which <- rep(1:72, each = 18)
best_mod <- dplyr::filter(all_liks, which == best_m)

@@

# Combine model likelihood table to benchmarks
bench_M3_joined <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikes_bench,
  by = "unit"
) %>%
  select(-ll_se) %>%
  mutate(unit = as.character(unit)) %>%
  pivot_longer(
    cols = -unit,
    names_to = "type",
    values_to = "ll"
  ) %>%
  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))

bench_diffs <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikes_bench,
  by = "unit"
) %>%
  rename(ll_mod3 = ll_est, ll_bench = ll) %>%
  mutate(unit = unlist(myLabeller(unit)),
         ll_diff = ll_mod3 - ll_bench) %>%
  arrange(ll_diff)

floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))
@

>>>>

<<DefineH3FitFunction>>=

# This function is how the model was fit before doing profiles. The commit id
# that this function was copied from, for reference, is 49b52e3 of haitipkg. 
fit_haiti3_si <- function(
    search1 = NULL,
    search2 = NULL,
    nsearches = 1,
    search_rho = FALSE,
    search_gamma = FALSE,
    search_hur = FALSE,
    search_Iinit = FALSE,
    start_date = "2010-11-20"
    ) {

  #
  ##
  ### Search1: Global search
  ##
  #

  if (is.null(search1$COOLING)) {
    search1$COOLING <- 0.5
  }

  if (is.null(search1$KEEP_LIK_MAT)) {
    search1$KEEP_LIK_MAT <- TRUE
  }

  if (nsearches >= 2 && is.null(search2$COOLING)) {
    search2$COOLING <- 0.5
  }

  if (nsearches >= 2 && is.null(search2$KEEP_LIK_MAT)) {
    search2$KEEP_LIK_MAT <- TRUE
  }

  # Create the model that will be fit to cholera incidence data
  h3_spat <- haitipkg::haiti3_spatPomp(start_date = start_date)

  # Create a list to save all of the results.
  results <- list()

  # Create vectors for the unit and shared parameters
  unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur")

  shared_param_names <- c(
    "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
    "epsilon", "k"
  )

  if (search_rho) {
    shared_param_names <- c(shared_param_names, "rho")
  }

  if (search_gamma) {
    shared_param_names <- c(shared_param_names, "gamma")
  }

  est_param_names <- c(
    unit_specific_names, shared_param_names
  )

  if (is.null(search1$RW_SD)) {
    stop("RW_SD must be supplied for search 1.")
  } else {
    chol_rw_1.sd <- search1$RW_SD
  }

  if (nsearches >= 2 && is.null(search2$RW_SD)) {
    stop("RW_SD must be supplied for search 2.")
  } else {
    chol_rw_2.sd <- search2$RW_SD
  }

  # Get lower bound for unit parameters (global search)
  min_val <- 1e-8
  unit_lb <- rep(c(1, 1e-9, 0, 0), each = 10)
  names(unit_lb) <- paste0(rep(unit_specific_names[1:4], each = 10), 1:10)

  # Get upper bound for unit parameters (global search)
  unit_ub <- rep(c(50, 1e-6, 0, 0), each = 10)
  names(unit_ub) <- paste0(rep(unit_specific_names[1:4], each = 10), 1:10)

  # Set unique upper-bounds for betaB, based on run_level_2 search results.
  unit_ub['betaB1'] <- 15
  unit_ub['betaB2'] <- 80
  unit_ub['betaB3'] <- 75
  unit_ub['betaB4'] <- 70
  unit_ub['betaB10'] <- 35
  unit_ub['betaB5'] <- 15
  unit_ub['betaB6'] <- 80
  unit_ub['betaB7'] <- 20
  unit_ub['betaB8'] <- 5
  unit_ub['betaB9'] <- 30
  unit_ub['foi_add10'] <- 2e-7
  unit_ub['foi_add3'] <- 2e-7
  unit_ub['foi_add4'] <- 2e-7
  unit_ub['foi_add6'] <- 5e-7
  unit_ub['foi_add9'] <- 5e-7

  if (search_hur) {
    unit_ub['aHur3'] <- 50
    unit_ub['aHur9'] <- 25
    unit_ub['hHur3'] <- 120
    unit_ub['hHur9'] <- 120

    unit_lb['aHur3'] <- 0.01
    unit_lb['aHur9'] <- 0.01
    unit_lb['hHur3'] <- 30
    unit_lb['hHur9'] <- 30
  }

  if (search_Iinit) {
    unit_lb['Iinit3'] <- 0.8 / 468301  # H3
    unit_lb['Iinit4'] <- 0.8 / 342525  # H4

    unit_ub['Iinit3']  <- 40 / 468301
    unit_ub['Iinit4']  <- 40 / 342525
  }

  # Get lower bound for shared parameters (global search)
  shared_lb <- c(
    25, 0.02, min_val, min_val,
    0.75, 0.005, 0.5, 1
  )

  if (search_rho) {
    shared_lb <- c(shared_lb, 1/8)
  }

  if (search_gamma) {
    shared_lb <- c(shared_lb, 365.25 / 5)
  }

  names(shared_lb) <- shared_param_names

  # Get upper bound for shared parameters (global search)
  shared_ub <- c(500, 0.8, 0.0001, 5, 1.75, 0.075, 0.99, 200)

  if (search_rho) {
    shared_ub <- c(shared_ub, 1/(0.2))
  }

  if (search_gamma) {
    shared_ub <- c(shared_ub, 365.25 / 2)
  }

  names(shared_ub) <- shared_param_names

  # Create data.frame with random unit parameters
  guesses_unit <- pomp::runif_design(
    lower = unit_lb,
    upper = unit_ub,
    nseq = search1$NREPS
  )

  # Create data.frame with random shared parameters
  guesses_shared <- pomp::runif_design(
    lower = shared_lb,
    upper = shared_ub,
    nseq = search1$NREPS
  )

  # Need to duplicate each of the shared parameter columns
  guesses_shared <- guesses_shared[, rep(1:length(shared_param_names), each = 10)]
  colnames(guesses_shared) <- paste0(rep(shared_param_names, each = 10), 1:10)

  # Combine the unit and shared parameters
  guesses <- cbind(guesses_shared, guesses_unit)

  # We need to add fixed parameters
  all_params <- coef(h3_spat)
  fixed_params <- all_params[!names(all_params) %in% colnames(guesses)]
  fixed_mat <- matrix(
    rep(fixed_params, search1$NREPS),
    byrow = TRUE, nrow = search1$NREPS
  )
  colnames(fixed_mat) <- names(all_params[!names(all_params) %in% colnames(guesses)])

  # Combine estimated and fixed parameters, and reorder based on original order.
  guesses_all <- cbind(guesses, fixed_mat)[, names(coef(h3_spat))]

  # Memory clean-up
  rm(guesses_unit, guesses_shared, fixed_mat, min_val, shared_lb,
     shared_ub, unit_lb, unit_ub,  all_params, fixed_params,
     est_param_names)
  gc()

  # It is assumed that a parallel backend has been registered prior to calling
  # this function.
  # doParallel::registerDoParallel(ncores)
  doRNG::registerDoRNG(2198635)

  t_global <- system.time(
    foreach(
      i = 1:search1$NREPS,
      .packages = c("spatPomp"),
      .combine = c
    ) %dopar% {
      r_params <- unlist(guesses_all[i, ])
      coef(h3_spat) <- r_params

      ibpf_out <- ibpf(
        h3_spat,
        Nbpf = search1$NBPF,
        Np = search1$NP,
        sharedParNames = shared_param_names,
        unitParNames = unit_specific_names,
        spat_regression = search1$SPAT_REGRESSION,
        rw.sd = chol_rw_1.sd,
        cooling.fraction.50 = search1$COOLING,
        block_size = 1
        # params = r_params  Not yet implemented in spatPomp
      )
    } -> Global_ibpf
  )

  rm(guesses_all)

  pfilterLikes <- data.frame(
    "ll" = rep(0, search1$NREPS_EVAL*search1$NREPS),
    "which" = rep(1:search1$NREPS, each = search1$NREPS_EVAL)
  )

  ibpf_params <- t(coef(Global_ibpf))

  for (i in 1:nrow(ibpf_params)) {
    for (sp in shared_param_names) {
      ibpf_params[i, paste0(sp, 1:10)] <- mean(ibpf_params[i, paste0(sp, 1:10)])
    }
  }

  t_global_bpf <- system.time(
    likMat <- foreach(  # Get log-likelihood for each unit and set of parameters, NREPS_EVAL times each
      i = 1:(search1$NREPS_EVAL*search1$NREPS), .combine = rbind, .packages = 'spatPomp'
    ) %dopar% {
      p3 <- ibpf_params[(i-1) %/% search1$NREPS_EVAL + 1, ]

      coef(h3_spat) <- p3
      apply(bpfilter(
        h3_spat, Np = search1$NP_EVAL,
        block_size = 1
      )@block.cond.loglik, 1, sum)
    }
  )

  # Condense unit likelihoods into model likelihood
  # pfilterLikes$ll <- apply(likMat, 1, sum)
  pfilterLikes$ll <- rowSums(likMat)  # Faster

  # Group by model parameter set.
  ibpf_logLik <- pfilterLikes |>
    dplyr::group_by(which) |>
    dplyr::summarize(logLik = logmeanexp(ll),
              se = logmeanexp(ll, se = TRUE)[2])

  search1_results <- list()
  search1_results$logLiks <- ibpf_logLik
  search1_results$params <- ibpf_params
  search1_results$ibpf_time <- t_global
  search1_results$bpf_time <- t_global_bpf

  if (isTRUE(search1$KEEP_LIK_MAT)) {
    search1_results$likMat <- likMat
  }

  if (isTRUE(search1$KEEP_TRACES)) {
    search1_results$traces <- lapply(Global_ibpf, function(x) x@traces[, c('loglik', names(chol_rw_1.sd@call)[-1])])
  }

  results$search1 <- search1_results

  if (nsearches == 1) {
    return(results)
  }

  rm(ibpf_logLik, guesses, likMat, pfilterLikes,
     Global_ibpf, chol_rw_1.sd, t_global, t_global_bpf,
     i, sp, ibpf_params)
  gc()

  #
  ##
  ### Search 2: Local Search
  ##
  #

  top_n_global <- order(-search1_results$logLiks$logLik)[1:search2$TOP_N]

  # Contains the best set of parameters, measure for the coupled model
  params <- search1_results$params[rep(top_n_global, each = search2$NREPS), ]

  all_params <- params

  rm(params)

  doRNG::registerDoRNG(327498615)

  t_ibpf_local <- system.time(
    foreach(
      i = 1:(search2$TOP_N * search2$NREPS),
      .packages = c("spatPomp"),
      .combine = c
    ) %dopar% {
      r_params <- all_params[i, ]
      coef(h3_spat) <- r_params

      ibpf_out <- ibpf(
        h3_spat,
        Nbpf = search2$NBPF,
        Np = search2$NP,
        sharedParNames = shared_param_names,
        unitParNames = unit_specific_names,
        spat_regression = search2$SPAT_REGRESSION,
        rw.sd = chol_rw_2.sd,
        cooling.fraction.50 = search2$COOLING,
        block_size = 1
        # params = r_params  Not yet implemented in spatPomp
      )
    } -> local_ibpf
  )

  rm(all_params)

  pfilterLikes <- data.frame(
    "ll" = NA_real_,
    "starting_set" = rep(top_n_global, each = search2$NREPS * search2$NREPS_EVAL),
    "which" = rep(1:(search2$NREPS * search2$TOP_N), each = search2$NREPS_EVAL)
  )

  ibpf_params <- t(coef(local_ibpf))

  for (i in 1:nrow(ibpf_params)) {
    for (sp in shared_param_names) {
      ibpf_params[i, paste0(sp, 1:10)] <- mean(ibpf_params[i, paste0(sp, 1:10)])
    }
  }

  t_local_bpf <- system.time(
    likMat <- foreach(  # Get log-likelihood for each unit and set of parameters, NREPS_EVAL times each
      i = 1:(search2$NREPS_EVAL*search2$NREPS*search2$TOP_N), .combine = rbind, .packages = 'spatPomp'
    ) %dopar% {
      # p3 <- coef(local_ibpf[[(i-1) %/% search2$NREPS_EVAL + 1]])
      p3 <- ibpf_params[(i-1) %/% search2$NREPS_EVAL + 1, ]
      coef(h3_spat) <- p3
      apply(bpfilter(
        h3_spat, Np = search2$NP_EVAL,
        block_size = 1
      )@block.cond.loglik, 1, sum)
    }
  )

  # Condense unit likelihoods into model likelihood
  pfilterLikes$ll <- rowSums(likMat)

  # Group by model parameter set.
  ibpf_logLik_temp <- pfilterLikes |>
    dplyr::group_by(which) |>
    dplyr::summarize(logLik = logmeanexp(ll),
                     se = logmeanexp(ll, se = TRUE)[2])

  ibpf_logLik <- pfilterLikes |>
    dplyr::group_by(which) |>
    dplyr::summarise(starting_set = dplyr::first(starting_set)) |>
    dplyr::right_join(
      y = ibpf_logLik_temp,
      by = "which"
    ) |>
    dplyr::select(which, starting_set, logLik, se)

  # ibpf_logLik$init_method <- (ibpf_logLik$which - 1) %/% (search2$TOP_N * search2$NREPS) + 1
  search2_results <- list()
  search2_results$logLiks <- ibpf_logLik
  search2_results$params <- ibpf_params
  search2_results$ibpf_time <- t_ibpf_local
  search2_results$bpf_time <- t_local_bpf

  if (isTRUE(search2$KEEP_LIK_MAT)) {
    search2_results$likMat <- likMat
  }

  if (isTRUE(search2$KEEP_TRACES)) {
    search2_results$traces <- lapply(local_ibpf, function(x) x@traces[, c('loglik', names(chol_rw_2.sd@call)[-1])])
  }

  results$search2 <- search2_results

  if (nsearches == 2) {
    return(results)
  }

  rm(
    search1, search1_results, top_n_global, t_ibpf_local, t_local_bpf,
    pfilterLikes, ibpf_logLik, chol_rw_2.sd, local_ibpf, likMat, ibpf_logLik_temp
  )

  gc()

  results
}
@

<<CalibrateOrig3>>=
unit_specific_names <- c("betaB", "foi_add")
shared_param_names <- c(
  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
  "epsilon", "k"
)
est_param_names <- c(
  unit_specific_names, shared_param_names
)

@@

haiti3_sub_fit <- bake(
  file = "../model3/si/h3_no_hurricane.rds", {
    fit_haiti3_si(
      nsearches = 2L,
      search1 = SEARCH1,
      search2 = SEARCH2,
      search_rho = FALSE,
      search_gamma = FALSE,
      search_Iinit = FALSE,
      search_hur = FALSE
    )
  }
)
@

<<LoadOrig3>>=

@@

# Perform block partical filter with fitted model, to get conditional
# log-likelihoods.
bpf_sub3_out <- bake(
  file = "../model3/si/noHur_cond_LL.rds", {

@@

# This result was used in the manuscript, so it doesn't need to be recalculated.
h3_final_mod_pars <- readRDS("../model3/run_level_3/h3_MLE_eval.rds") %>%
  slice_max(logLik, n = 1) %>%
  select(-(job.id:logLik_se)) %>%
  unlist()

# h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")

# Save best_m as index for the model parameters with the best likelihood.
# Then, save the model parameter values and the likelihood evaluations
# best_m <- which.max(h3_final_results$search2$logLiks$logLik)
# best_params <- h3_final_results$search2$params[best_m, ]
# all_liks <- as.data.frame(h3_final_results$search2$likMat)
# all_liks$which <- rep(1:72, each = 18)
# best_mod <- dplyr::filter(all_liks, which == best_m)

# Change model parameters to the best set of parameters from previous
# search
coef(h3)[names(h3_final_mod_pars)] <- h3_final_mod_pars

registerDoRNG(731955)  # for modular, reproducible results

# Perform block partical filter with fitted model, to get conditional
# log-likelihoods.
bpf_final3_out <- bake(
  file = "../model3/si/h3_cond_LL.rds", {

    results$block.cond.loglik <- bpf_out@block.cond.loglik
    results$times <- bpf_out@times  # Saved for convenience
    results
  },
  seed = 432235
)

@@

# # A tibble: 10 × 3
 #    unit ll_est ll_se
 #   <dbl>  <dbl> <dbl>
 # 1     1 -2122. 0.221
 # 2     2 -2135. 0.419
 # 3     3 -1529. 0.882
 # 4     4 -1134. 0.501
 # 5     5 -1902. 0.213
 # 6     6 -1423. 0.303
 # 7     7 -1615. 0.633

# Combine model likelihood table to benchmarks
bench_M3_joined <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikes_bench,
  by = "unit"
) %>%
  select(-se) %>%
  mutate(unit = as.character(unit)) %>%
  pivot_longer(
    cols = -unit,
    names_to = "type",
    values_to = "ll"
  ) %>%
  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))

bench_diffs <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikes_bench,
  by = "unit"
) %>%
  rename(ll_mod3 = est, ll_bench = ll) %>%
  mutate(unit = unlist(myLabeller(unit)),
         ll_diff = ll_mod3 - ll_bench) %>%
  arrange(ll_diff)

# floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))
@
